---
title: "Analyzing the Relationship between Sentiment and Language in English-Tamil Code-Switching"
author: "Paul Bontempo"
date: "`12/1/2025"
output:
  html_document:
    toc: true
    toc_float: true
  word_document: default
fontsize: 11pt
geometry: margin=1in
---

# 1. INTRODUCTION

This project investigates the relationship between sentiment and language choice in code-switched English-Tamil text from YouTube comments. We hypothesize that sentiment predicts both the proportion of English used and the frequency of language switches within utterances.

**Research Questions:**
1. Does sentiment category predict the proportion of English vs. Tamil in code-switched utterances?
2. Does sentiment category predict the frequency of language switches within utterances?

**Hypothesis:** Mixed_feelings utterances will show lower English proportions than Positive or Negative utterances, reflecting English's role as a prestige/academic language versus Tamil's role as an emotional/home language.

# 2. DATA IMPORT AND PREPARATION

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(car)

# Import data
cs_data <- read.csv("codeswitching_data.csv", stringsAsFactors = FALSE)

# Convert sentiment to factor with Positive as reference level
cs_data$sentiment <- factor(cs_data$sentiment, 
                            levels = c("Positive", "Mixed_feelings", "Negative", "unknown"))

# Check structure
str(cs_data)
summary(cs_data)
```

## 2.1 Data Overview

```{r data_overview}
# Sentiment distribution
table(cs_data$sentiment)

# Overall proportions
cat("Overall NA proportion:", mean(cs_data$na_proportion), "\n")
cat("Overall English proportion:", mean(cs_data$en_proportion), "\n")
cat("Overall Tamil proportion:", mean(cs_data$ta_proportion), "\n")
cat("Overall switches per sentence:", mean(cs_data$num_switches), "\n")
```

## 2.2 Filter to Main Sentiment Categories

Focus on the three clear sentiment categories for primary analysis:

```{r filter_main}
# Create filtered dataset for main analysis
cs_main <- cs_data %>%
  filter(sentiment %in% c("Positive", "Mixed_feelings", "Negative"))

# Re-level factor to drop unused level
cs_main$sentiment <- droplevels(cs_main$sentiment)

# Summary of filtered data
cat("Sentences in main analysis:", nrow(cs_main), "\n")
table(cs_main$sentiment)
```

# 3. EXPLORATORY DATA ANALYSIS

## 3.1 Distribution of English Proportion

```{r hist_en_prop, fig.width=8, fig.height=5}
ggplot(cs_main, aes(x = en_proportion)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "black") +
  facet_wrap(~sentiment) +
  labs(title = "Distribution of English Proportion by Sentiment",
       x = "English Proportion", y = "Count") +
  theme_minimal()
```

## 3.2 English Proportion by Sentiment

```{r boxplot_en_prop, fig.width=8, fig.height=5}
ggplot(cs_main, aes(x = sentiment, y = en_proportion, fill = sentiment)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  labs(title = "English Proportion by Sentiment",
       x = "Sentiment", y = "English Proportion") +
  theme_minimal() +
  theme(legend.position = "none")

# Summary statistics
cs_main %>%
  group_by(sentiment) %>%
  summarise(
    n = n(),
    mean_en = mean(en_proportion),
    sd_en = sd(en_proportion),
    median_en = median(en_proportion)
  )
```

## 3.3 Number of Switches by Sentiment

```{r boxplot_switches, fig.width=8, fig.height=5}
ggplot(cs_main, aes(x = sentiment, y = num_switches, fill = sentiment)) +
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  labs(title = "Language Switches per Sentence by Sentiment",
       x = "Sentiment", y = "Number of Switches") +
  theme_minimal() +
  theme(legend.position = "none")

# Summary statistics
cs_main %>%
  group_by(sentiment) %>%
  summarise(
    n = n(),
    mean_switches = mean(num_switches),
    sd_switches = sd(num_switches),
    median_switches = median(num_switches)
  )
```

# 4. ANALYSIS 1: ENGLISH PROPORTION ~ SENTIMENT

## 4.1 Simple Linear Model

```{r model1_simple}
# Model 1: English proportion predicted by sentiment
model1 <- lm(en_proportion ~ sentiment, data = cs_main)
summary(model1)
```

## 4.2 Model Diagnostics

```{r model1_diagnostics, fig.width=10, fig.height=8}
par(mfrow = c(2, 2))
plot(model1)
par(mfrow = c(1, 1))
```

## 4.3 Check for Heteroscedasticity

```{r model1_heteroscedasticity}
# Levene's test for homogeneity of variance
leveneTest(en_proportion ~ sentiment, data = cs_main)
```

## 4.4 Coefficient Plot with Confidence Intervals

```{r model1_coef_plot, fig.width=8, fig.height=5}
# Extract coefficients
model1_coefs <- tidy(model1, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(estimate) %>%
  mutate(term = factor(term, levels = term))

# Dot-and-whiskers plot
ggplot(model1_coefs, aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Sentiment on English Proportion",
       subtitle = "Reference: Positive sentiment",
       x = "Predictor", y = "Coefficient Estimate (95% CI)") +
  theme_minimal()
```

## 4.5 Interpretation

```{r model1_interpretation}
# Model fit
cat("R-squared:", summary(model1)$r.squared, "\n")
cat("Adjusted R-squared:", summary(model1)$adj.r.squared, "\n\n")

# Predicted means
predicted_means <- cs_main %>%
  group_by(sentiment) %>%
  summarise(observed_mean = mean(en_proportion)) %>%
  mutate(predicted_mean = predict(model1, newdata = data.frame(sentiment = sentiment)))

print(predicted_means)
```

# 5. ANALYSIS 2: SWITCHES ~ SENTIMENT

## 5.1 Simple Linear Model for Count Data

```{r model2_simple}
# Model 2: Number of switches predicted by sentiment
model2 <- lm(num_switches ~ sentiment, data = cs_main)
summary(model2)
```

## 5.2 Model Diagnostics

```{r model2_diagnostics, fig.width=10, fig.height=8}
par(mfrow = c(2, 2))
plot(model2)
par(mfrow = c(1, 1))
```

## 5.3 Coefficient Plot

```{r model2_coef_plot, fig.width=8, fig.height=5}
# Extract coefficients
model2_coefs <- tidy(model2, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(estimate) %>%
  mutate(term = factor(term, levels = term))

# Dot-and-whiskers plot
ggplot(model2_coefs, aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Sentiment on Language Switches",
       subtitle = "Reference: Positive sentiment",
       x = "Predictor", y = "Coefficient Estimate (95% CI)") +
  theme_minimal()
```

# 6. EXTENDED ANALYSIS: INTERACTION WITH SENTENCE LENGTH

## 6.1 Model with Interaction Term

```{r model3_interaction}
# Model 3: Switches predicted by sentiment and sentence length with interaction
model3 <- lm(num_switches ~ sentiment * sentence_length, data = cs_main)
summary(model3)
```

## 6.2 Model Comparison

```{r model_comparison}
# Compare models with ANOVA
anova(model2, model3)

# Compare R-squared values
cat("Model 2 (no interaction) R-squared:", summary(model2)$r.squared, "\n")
cat("Model 3 (with interaction) R-squared:", summary(model3)$r.squared, "\n")
```

## 6.3 Visualization of Interaction

```{r interaction_plot, fig.width=10, fig.height=6}
# Create predicted values for visualization
pred_data <- expand.grid(
  sentiment = levels(cs_main$sentiment),
  sentence_length = seq(min(cs_main$sentence_length), 
                        max(cs_main$sentence_length), 
                        length.out = 50)
)

pred_data$predicted_switches <- predict(model3, newdata = pred_data)

ggplot(pred_data, aes(x = sentence_length, y = predicted_switches, color = sentiment)) +
  geom_line(size = 1) +
  labs(title = "Predicted Language Switches by Sentiment and Sentence Length",
       x = "Sentence Length (tokens)", y = "Predicted Number of Switches",
       color = "Sentiment") +
  theme_minimal()
```

# 7. RESULTS SUMMARY

```{r results_summary}
cat("=== MAIN FINDINGS ===\n\n")

cat("Analysis 1: English Proportion ~ Sentiment\n")
cat("Reference level: Positive sentiment\n")
cat("Model R-squared:", round(summary(model1)$r.squared, 4), "\n\n")

model1_summary <- tidy(model1) %>% filter(term != "(Intercept)")
for(i in 1:nrow(model1_summary)) {
  cat(model1_summary$term[i], ":\n")
  cat("  Estimate:", round(model1_summary$estimate[i], 4), "\n")
  cat("  p-value:", format.pval(model1_summary$p.value[i]), "\n\n")
}

cat("\nAnalysis 2: Language Switches ~ Sentiment\n")
cat("Reference level: Positive sentiment\n")
cat("Model R-squared:", round(summary(model2)$r.squared, 4), "\n\n")

model2_summary <- tidy(model2) %>% filter(term != "(Intercept)")
for(i in 1:nrow(model2_summary)) {
  cat(model2_summary$term[i], ":\n")
  cat("  Estimate:", round(model2_summary$estimate[i], 4), "\n")
  cat("  p-value:", format.pval(model2_summary$p.value[i]), "\n\n")
}
```

# 8. DISCUSSION

[TO BE COMPLETED AFTER RUNNING ANALYSIS]

**Key findings:**
- [Describe whether Mixed_feelings has lower English proportion than Positive]
- [Describe relationship between sentiment and language switches]
- [Interpret interaction effects if significant]

**Limitations:**
- [List any limitations discovered during analysis]

**Future directions:**
- [Suggest refinements or extensions]