---
title: "Code-Switching and Sentiment in English-Tamil Bilingual Text"
author: "Paul Bontempo"
date: "`r Sys.Date()`"
output:
  word_document
fontsize: 11pt
geometry: margin=1in
---

# INTRODUCTION

This project investigates the relationship between sentiment and language choice in code-switched English-Tamil text from YouTube comments. We hypothesize that sentiment predicts both the proportion of English used and the frequency of language switches within utterances.

**Research Questions:**
1. Does sentiment category predict the proportion of English vs. Tamil in code-switched utterances?
2. Does sentiment category predict the frequency of language switches within utterances?

**Proportion hypotheses**
H0: Utterance sentiment has no effect on proportion of embedded language to matrix language
H1: Utterance sentiment predicts the proportion of embedded language to matrix language
**Frequency hypotheses**
H0: Utterance sentiment has no effect on the frequency of language switches
H1: Utterance sentiment predicts the frequency of language switches

# DATA IMPORT AND PREPARATION

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(car)

# Import data
cs_data <- read.csv("entacs_data_complete.csv", stringsAsFactors = FALSE)

# Convert sentiment to factor with Positive as reference level
cs_data$sentiment <- factor(cs_data$sentiment, 
                            levels = c("Positive", "Mixed_feelings", "Negative", "unknown"))

# Check structure
summary(cs_data)
```

## Data Overview

```{r data_overview}
# Sentiment distribution
table(cs_data$sentiment)

# Overall proportions
cat("Overall NA proportion:", mean(cs_data$na_proportion), "\n")
cat("Overall English proportion:", mean(cs_data$en_proportion), "\n")
cat("Overall Tamil proportion:", mean(cs_data$ta_proportion), "\n")
cat("Overall switches per sentence:", mean(cs_data$num_switches), "\n")
```

## Filter to Main Sentiment Categories

Focus on the three clear sentiment categories for primary analysis:

```{r filter_main}
# Create filtered dataset for main analysis
cs_main <- cs_data %>%
  filter(sentiment %in% c("Positive", "Mixed_feelings", "Negative"))

# Re-level factor to drop unused level
cs_main$sentiment <- droplevels(cs_main$sentiment)

# Summary of filtered data
cat("Sentences in main analysis:", nrow(cs_main), "\n")
table(cs_main$sentiment)
```

## Distribution of English Proportion

```{r hist_en_prop, fig.width=8, fig.height=5}
ggplot(cs_main, aes(x = en_proportion)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "black") +
  facet_wrap(~sentiment) +
  labs(title = "Distribution of English Proportion by Sentiment",
       x = "English Proportion", y = "Count") +
  theme_minimal()
```

## Boxplot: English Proportion by Sentiment

```{r boxplot_en_prop, fig.width=8, fig.height=5}
ggplot(cs_main, aes(x = sentiment, y = en_proportion, fill = sentiment)) +
  geom_boxplot(outlier.shape = 1) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  ylim(0, 1) +  # Show full scale from 0 to 1
  labs(title = "English Proportion by Sentiment",
       x = "Sentiment", y = "English Proportion") +
  theme_minimal() +
  theme(legend.position = "none")

# Summary statistics
cs_main %>%
  group_by(sentiment) %>%
  summarise(
    n = n(),
    mean_en = mean(en_proportion),
    sd_en = sd(en_proportion),
    median_en = median(en_proportion)
  )
```

## Boxplot: Number of Switches by Sentiment

```{r boxplot_switches, fig.width=8, fig.height=5}
ggplot(cs_main, aes(x = sentiment, y = num_switches, fill = sentiment)) +
  geom_boxplot(outlier.shape = 1) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  coord_cartesian(ylim = c(0, max(cs_main$num_switches))) +  # Show all data including outliers
  labs(title = "Language Switches per Utterance by Sentiment",
       x = "Sentiment", y = "Number of Switches") +
  theme_minimal() +
  theme(legend.position = "none")

# Summary statistics
cs_main %>%
  group_by(sentiment) %>%
  summarise(
    n = n(),
    mean_switches = mean(num_switches),
    sd_switches = sd(num_switches),
    median_switches = median(num_switches)
  )
```

# EXPERIMENT 1: LANGUAGE AS A FUNCTION OF SENTIMENT

## 1a: Simple Linear Model

```{r model1_simple}
# Model 1a: English proportion predicted by sentiment
model1a <- lm(en_proportion ~ sentiment, data = cs_main)
summary(model1a)
```

## Q-Q Plot (Simple Model)

```{r model1a_qq_plot, fig.width=6, fig.height=6}
# Q-Q plot to check normality of residuals
qqnorm(resid(model1a), main = "Q-Q Plot Residuals (Model 1a)")
qqline(resid(model1a), col = "red")
```

## Coefficient Plot w/ Confidence Intervals (Simple Model)

```{r model1a_coef_plot, fig.width=8, fig.height=5}
# Extract coefficients
model1a_coefs <- tidy(model1a, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(estimate) %>%
  mutate(term = factor(term, levels = term))

# Dot-and-whiskers plot
ggplot(model1a_coefs, aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Sentiment on English Proportion",
       subtitle = "(Model 1a; Reference Level: Positive)",
       x = "Predictor", y = "Coefficient Estimate (95% CI)") +
  theme_minimal()
```

## Model Fit (Simple Model)

```{r model1a_interpretation}
# Model fit
cat("R-squared:", summary(model1a)$r.squared, "\n")
cat("Adjusted R-squared:", summary(model1a)$adj.r.squared, "\n\n")
```

## 1b: Interaction Model Proportion ~ Sentiment * Sentence Length

```{r model1b_interaction}
# Model 1b: English proportion predicted by sentiment, sentence length, and their interaction
model1b <- lm(en_proportion ~ sentiment * sentence_length, data = cs_main)
summary(model1b)
```

## Q-Q Plot (Interaction Model)

```{r model1b_qq_plot, fig.width=6, fig.height=6}
# Q-Q plot for model with interaction
qqnorm(resid(model1b), main = "Q-Q Plot Residuals (Model 1b with Interaction)")
qqline(resid(model1b), col = "red")
```

## Coefficient Plot w/ Confidence Intervals (Interaction Model)

```{r model1b_coef_plot, fig.width=8, fig.height=6}
# Extract coefficients from interaction model
model1b_coefs <- tidy(model1b, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(estimate) %>%
  mutate(term = factor(term, levels = term))

# Dot-and-whiskers plot
ggplot(model1b_coefs, aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Sentiment and Length on English Proportion",
       subtitle = "(Model 1b; Reference Level: Positive)",
       x = "Predictor", y = "Coefficient Estimate (95% CI)") +
  theme_minimal()
```

## 1c: Model Comparison

```{r model1_comparison}
# Compare models with ANOVA
anova(model1a, model1b)

# Compare R-squared values
cat("Model 1a R-squared:", summary(model1a)$r.squared, "\n")
cat("Model 1b (with interaction) R-squared:", summary(model1b)$r.squared, "\n")
```

## Interaction Visualization

```{r interaction_en_plot, fig.width=10, fig.height=6}
# Plot actual data points with fitted regression lines
ggplot(cs_main, aes(x = sentence_length, y = en_proportion, color = sentiment)) +
  geom_point(alpha = 0.3, size = 0.8) +  # Actual data points
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +  # Fitted lines with confidence bands
  ylim(0, 1) +
  labs(title = "English Proportion by Sentiment and Sentence Length",
       subtitle = "Points show actual data; lines show linear model fits",
       x = "Sentence Length (tokens)", 
       y = "English Proportion",
       color = "Sentiment") +
  theme_minimal()
```

# EXPERIMENT 2: SWITCHES AS FUNCTION OF SENTIMENT

## 2a: Simple Linear Model

```{r model2a_simple}
# Model 2a: Number of language switches per utterance predicted by sentiment
model2a <- lm(num_switches ~ sentiment, data = cs_main)
summary(model2a)
```

## Q-Q Plot (Simple Model)

```{r model2a_qq_plot, fig.width=6, fig.height=6}
# Q-Q plot to check normality of residuals
qqnorm(resid(model2a), main = "Q-Q Plot Residuals (Model 2a)")
qqline(resid(model2a), col = "red")
```

## Coefficient Plot (Simple Model)

```{r model2a_coef_plot, fig.width=8, fig.height=5}
# Extract coefficients
model2a_coefs <- tidy(model2a, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(estimate) %>%
  mutate(term = factor(term, levels = term))

# Dot-and-whiskers plot
ggplot(model2a_coefs, aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Sentiment on Language Switches",
       subtitle = "(Model 2a; Reference: Positive sentiment)",
       x = "Predictor", y = "Coefficient Estimate (95% CI)") +
  theme_minimal()
```

## Model Fit (Simple Model)

```{r model2a_interpretation}
# Model fit
cat("R-squared:", summary(model2a)$r.squared, "\n")
cat("Adjusted R-squared:", summary(model2a)$adj.r.squared, "\n\n")
```

## 2b: Interaction Model Language Switches ~ Sentiment * Sentence Length

```{r model2b_interaction}
# Model 2b: Switches predicted by sentiment, sentence length, and their interaction
model2b <- lm(num_switches ~ sentiment * sentence_length, data = cs_main)
summary(model2b)
```

## Q-Q Plot (Interaction Model)

```{r model2b_qq_plot, fig.width=6, fig.height=6}
# Q-Q plot for model with interaction term
qqnorm(resid(model2b), main = "Q-Q Plot Residuals (Model 2b with Interaction)")
qqline(resid(model2b), col = "red")
```

## Coefficient Plot (Interaction Model)

```{r model2b_coef_plot, fig.width=8, fig.height=5}
# Extract coefficients
model2b_coefs <- tidy(model2b, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(estimate) %>%
  mutate(term = factor(term, levels = term))

# Dot-and-whiskers plot
ggplot(model2b_coefs, aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Effect of Sentiment and Length on Language Switches",
       subtitle = "(Model 2b; Reference Level: Positive)",
       x = "Predictor", y = "Coefficient Estimate (95% CI)") +
  theme_minimal()
```

## 2c: Model Comparison

```{r model2_comparison}
# Compare models with ANOVA
anova(model2a, model2b)

# Compare R-squared values
cat("Model 2a R-squared:", summary(model2a)$r.squared, "\n")
cat("Model 2b (with interaction) R-squared:", summary(model2b)$r.squared, "\n")
```

## Visualization of Interaction

```{r interaction_plot, fig.width=10, fig.height=6}
# Plot actual data points with fitted regression lines
ggplot(cs_main, aes(x = sentence_length, y = num_switches, color = sentiment)) +
  geom_point(alpha = 0.3, size = 0.8) +  # Actual data points
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +  # Fitted lines with confidence bands
  labs(title = "Language Switches by Sentiment and Sentence Length",
       subtitle = "Points show actual data; lines show linear model fits",
       x = "Sentence Length (tokens)", 
       y = "Number of Language Switches",
       color = "Sentiment") +
  theme_minimal()
```


